---
layout: post
title: Variational Inference for MRP with Reliable Posterior Distributions
subtitle: Part 4- Some theory on why VI is hard
date: 2023-05-02
draft: True
categories:
- MRP
- Variational Inference
---

This is section 4 in my series on using Variational Inference to speed up relatively complex Bayesian models like Multilevel Regression and Poststratification without the approximation being of disastrously poor quality. 

The general structure for this post and the one before and after it will be to describe a problem with VI, and then describe how that problem can be fixed to some degree. Collectively, all the small improvements in these three posts will go a long way towards more robust variational inference. I'll also have a grab bag at the end of other interesting ideas from the literature I think are cool, but maybe not as important or interesting to me as the 3 below.

In the [last post](https://andytimm.github.io/posts/Variational%20MRP%20Pt3/variational_mrp_3.html) we took a look at how our ELBO objective requires  specific version of KL Divergence (the "Exclusive" formulation of KLD), and saw that it encoded a preference for a certain type of solution to the VI problem. Then we looked at CUBO and CHIVI, an alternative bound and algorithm that avoid this problem, often leading to a more useful posterior distribution by pursuing a more "inclusive" solution.

The rough plan for the series is as follows:

1.  Introducing the Problem- Why is VI useful, why VI can produce spherical cows
2.  How far does iteration on classic VI algorithms like mean-field and full-rank get us?
3.  Problem 1: KL-D prefers exclusive solutions; are there alternatives?
4. **(This post)** Problem 2: Not all VI samples are of equal utility; can we weight them cleverly?
5. Problem 3: How can we know when VI is wrong? Are there useful error bounds?
5. Better grounded diagnostics and workflow
6. Seeing if some more sophisticated techniques like normalizing flows add much

# Not all samples are equally good

So we've made an approximation $q(x)$ that's cheap to sample from, and is somewhat
close to $p(x)$, our true posterior. The way to improve the approximation we've
focused on so far is to just go back to the start and make $q(x)$ better; for example,
through changing up the variational family, or to switching to a different
optimization objective like the CUBO. That's one solution that's often necessary,
but can we work with a particular $q(x)$ we have and make better use of the parts of it that are the closest to being right?

... Phased this way, this sounds a lot like importance sampling. If you haven't seen them before,
an importance sampling estimator allows us to take draws from a (preferably) easy to sample
from distribution[^7] and reweight the samples to look more like our true target
distribution. The weight $w_i$ for each sample $i$ take form:

$$
w_i = \frac{p(x_i)}{q(x_i)}
$$
Before you get worried that we don't have $p(x_i)$ because of the normalizing constant like every time we talk about having $p(x)$ in this series, there's a clever estimator that
"self-normalizes" such that this can be a reasonable strategy. Intuitively, we're
just placing more weight on samples in the support of $p(x)$.

This footenote[^8] has a selection of some of my favorite resources for learning more or refreshing your memory about importance sampling, but for the main discussion let me pull out some
particularly important sub-problems to solve in making a good importance sampling estimator, and good important sampling estimator for VI. 

First, our choice of the "proposal" distribution we're reweighting to be more like
$p(x)$ matters for making this process practically feasible. We need the proposal
distribution to be close enough to $p(x)$ that a realistic number of the draws
get non-negligible weights.
It might be true that we could draw proposals from a big $N$ dimensional uniform
distribution for every problem, but if we want to be done sampling enough this century
we need to at least get fairly close with our initial $q(x)$.

A second, but related problem is that it's quite common for the unmodified
importance sampling estimator to have some weights which are orders and orders
of magnitude higher than the average weight, blowing up the variance of the estimator.
Dan Simpson's slides I linked above has an instructive example with not too
weird $p(x)$ and $q(x)$'s that has a max weight ~1.4 million (!) times the average.
If that happens, our estimator will essentially ignore most samples without gigantic
weights, and it'll take ages for that estimator to tell us anything remotely
reliable. 

So with those points we need to address, here are the next topics in this post:

1. Importance Weighted Variational Inference
2. Robust importance sampling with built in diagnostics via PSIS
3. Combining multiple proposal distributions via Multiple Importance Sampling


## Importance Weighted Variational Inference

The actual use of Importance Weighting for VI in it's simplest form is
pretty intuitive (draw samples from $q(x)$, weight them...), but let's derive the
new Importance Weighted Variational Inference (IWVI) estimator since some nice
intuition will come with it.

We'll aim to show that we can get a tighter ELBO by using importance weights. This
type of tighter ELBO was first shown by [Burda et Al. (2015)](https://arxiv.org/abs/1509.00519)
in the context of Variational Autoencoders after which is was fairly clear this could apply
to variational inference, but [Domke and Sheldon (2018)](https://arxiv.org/abs/1808.09034)
fleshed out some details of that extension- I'll be explaining some of the latter group's main
results first.

To start, imagine a random variable $R$, such that $\mathbb{E}{R} = p(x)$, which
we'll think of as a estimator of p(x). Then by Jensen's Inequality:

$$
logp(x) = \mathbb{E}logR + \mathbb{E}log\frac{p(x)}{R}
$$

The first term is the bound, which will be tighter if $R$ is highly concentrated. 

This is a more general form of the ELBO; we can make it quite familiar looking by
having our R above be:

$$
R = \frac{p(z,x)}{q(z)}, z \sim q
$$

The reason pointing out this fairly simple generalization is helpful is that it
frames how to tighten our ELBO bound on $logp(x)$ via alternative estimators $R$.

By drawing $M$ samples and averaging them as in importance sampling, we get:

$$
R_M = \frac{1}{M}\sum_{m=1}^{M}\frac{p(z_m,x)}{q(z_m)}, z_m \sim q
$$
From there, we can derive a tighter bound on $logp(x)$, referred to as the IW-ELBO:

$$
IW-ELBO_M[q(z)||p(z,x)] := \mathbb{E}_{q(z_{1:M})}log\frac{1}{M} \sum_{m=1}^{M}\frac{p(z_m,x)}{q(z_m)}
$$
Where we're using the $1:M$ as a shorthand for eg $q(z_{1:M}) = q(z_1)...q(z_M)$.


It's worth noting that the last few lines don't specify a particular form of importance
sampling- we're getting the tighter theoretical bounding behavior from the averaging of samples
from $q$. We'll see a particularly good form of importance sampling with desirable
practical properties in a moment.

### How does IW-ELBO change the VI problem conceptually?

The tighter bound is nice, but importance sampling also has the side effect (done right, side benefit)
of modifying our incentives in choosing a variational family. To see what I mean,
we can re-use the example distributions from last post we used to build intuition for KL Divergence, where red was the true distribution,
and green were our potential approximations. If we're not going to draw multiple samples and weight them, it makes sense to choose something like the first plot below. Every draw in the middle of the two target modes
is expensive per our ELBO objective, so better to choose a mode.

```{r, echo= FALSE, warning=FALSE, output = FALSE}
library(ggplot2)
library(gridExtra)
library(tidyverse)
```


```{r,echo=F}

mixture <- data.frame(normals = c(rnorm(1000,3,1),rnorm(1000,15,2)),
                      mode_seeking_kl = rnorm(2000,3.5,2),
                      mean_seeking_kl = rnorm(2000,9,4))
```


```{r}


rkl_plot <- mixture %>% ggplot(aes(x = normals)) +
  geom_density(aes(x = normals), color = "red") +
  geom_density(aes(x = mode_seeking_kl), color = "green") + ggtitle("Without weighting, we prefer to capture a mode") +
  xlab("")

fkl_plot <- mixture %>% ggplot(aes(x = normals)) +
  geom_density(aes(x = normals), color = "red") +
  geom_density(aes(x = mean_seeking_kl), color = "green") + ggtitle("With importance sampling, weights allow us to prefer coverage") +
  xlab("")

grid.arrange(rkl_plot,fkl_plot)
```

If we can use importance samples though, quite the opposite is be true! Note that we're
still using the ELBO, a reverse-KL based metric- that hasn't changed. What has changed is our
ability to mitigate the objective costs of those samples between the two extremes.
Points outside the two target modes will get lower importance weights, and points within
the modes will get higher ones, so as long as we're covering the modes with some reasonable amount of probability mass, and drawing enough samples we can actually
do better with the distribution centered between the modes. 

We can also imagine how varying the number of samples might effect the point
from the last paragraph. Between $M=1$ and "enough draws to get all the benefits of IS",
we can imagine there's a slow transition from "just stick with 1 mode" and "go with IS".
So it seems like we should be worried about getting the number of samples right, but
fortunately as we'll see in the next section there are great rules of thumb in
some variants of IS. We'll still need to bear the cost of sampling (which gets higher as $q(x)$ becomes "further" from $p(x)$, as we'll need more samples to weight into a good approximation), but
the cost of sampling for most VI implementations will often be pretty manageable
if our proposal distribution is somewhat close to $p(x)$.

Another way to think about how importance sampling changes our task with variational
inference is to think about what sorts of distributions make sense to have as
our variational family, and even which objective might be better given IS. On choice
of a variational family, if we're aiming for coverage, moving towards thicker-tailed distributions like t distributions makes a lot of sense. While we explored
the IW-ELBO above to build intuition, there's no reason not to apply VI to the CUBO
 and thus CHIVI- this also naturally produces nicely overdispersed distributions which can
 be importance sampled closer to the true $p(x)$. This idea of aiming for a
 wide proposal to sample from is referred to in the importance sampling literature (eg [Owen, 2013](https://artowen.su.domains/mc/)) as "defensive sampling", with [Domke and Sheldon (2018)](https://arxiv.org/abs/1808.09034) exploring the VI connection more fully. For intuition, by ensuring most of p(x) is covered by some reasonable mass makes it easier to efficiently
 get draws that can be weighted into a final posterior, even if the unweighted
 posterior might be too wide.

## Solving our IS problems with Pareto-Smoothed Importance Sampling

As we've been talking about importance sampling, we've been leaving some of the
messier details aside (how many samples to draw, how to deal with the cases when some
of the weights get huge, how to know when our proposal distribution is "close" enough).

While the Importance Sampling Literature is huge and there are a lot of possible
solutions here, I'll next introduce [Vehtari et Al. (2015)](https://arxiv.org/abs/1507.02646)'s Pareto-Smoothed Importance Sampling. I'm a huge fan of this paper, and it's
amongst the coolest things I got to study in grad school. 

The core idea here as I started to describe above is that.

## Multiple Proposal Distributions with Multiple Importance Sampling

# Can we bound error in terms of ELBO or CUBO?

## Wasserstein Bounds

# Conclusions + Bonus Context


[^7]: we'll call it q(x) here to make the application super clear, but often
I see the "proposal" distribution called f(x) and the the distribution we want
to approximate called g(x). 
[^8]: If you're looking to learn about importance sampling for the first time,
a great place to start is Ben Lambert's video introductions to the basic idea:
[video 1](https://www.youtube.com/watch?v=V8f8ueBc9sY), and [video 2](https://www.youtube.com/watch?v=F5PdIQxMA28). For building more intuition about
why we need all these variance reducing modifications to general IS, Dan Simpson
has some great [slides](https://dpsimpson.github.io/pages/talks/Importance_sampling_unsw_2019.pdf) which have a side benefit of being hilarious. Those slides will mention a lot of the books/papers
I find most instructive, but it's worth calling out especially Vehtari et Al's Pareto Smoothed
Importance Sampling [paper](https://arxiv.org/abs/1507.02646) as particularly
well written and paradigm shaping. Finally, Elvira et Al's (2019) Multiple Importance Sampling
[paper](https://projecteuclid.org/journals/statistical-science/volume-34/issue-1/Generalized-Multiple-Importance-Sampling/10.1214/18-STS668.full) is the most thorough I know, but isn't particularly approachable. Instead, for MIS I'd recommend starting with the first few minutes of [this talk](https://www.youtube.com/watch?v=dxFSwplfdpk) (although the main topic of their talk is less relevant, the visualizations are super helpful), and the first ~8 pages of [this paper](https://arxiv.org/pdf/2102.05407.pdf), also by Elvira et Al. (2021) (I especially like
that it spends a bit more time on notation; since multiple importance sampling comes from/comes up
in computer graphics, the notational choices sometimes feel a bit annoying to me). Finally,
the [original MIS paper itself](https://dl.acm.org/doi/10.1145/218380.218498), Veach & Guibas (1995) is quite readable, but requires
a bit of reading around or reading into computer graphics to grok their examples
and notational choices. 